Time and space specifications for stencils

### 2016-03-03 paper.tex
more intro; towards a definition of sub stencils

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{color}
\usepackage{graphics}
\usepackage{fancyvrb}
\CustomVerbatimEnvironment{SpecVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.65cm,
xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.97}
\CustomVerbatimEnvironment{ExmVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.65cm,
xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.97}
\CustomVerbatimEnvironment{IVerbatim}{Verbatim}{fontsize=\relsize{-1},xleftmargin=0.65cm,
xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.97}

@@

@@

\title{Abstract Shape Specifications for Stencil Verification}
\authorinfo{}{}

@@

\section{Specification language}

\begin{align*}
spec ::= & \; S  \; \texttt{,} \; spec \, \mid \, S \\[0.5em]
%
S ::= &  \; \texttt{forward} \; \texttt{depth=}\textit{depth} \; \texttt{dims=}\textit{dims} \\
& \; \mid \texttt{backward} \; \texttt{depth=}\textit{depth} \; \texttt{dims=}\textit{dims} \\
& \; \mid \texttt{centered} \; \texttt{depth=}\textit{depth} \; \texttt{dims=}\textit{dims} \\
& \; \mid \texttt{unspecified} \; \texttt{dims=}[\textit{dims}] \\
& \; \mid \texttt{read-once} \; S \\[0.5em]
\textit{depth} ::= & \; \mathbb{N} \\
\textit{dims}  ::= & \; \textit{var} \; \texttt{,} \; \textit{dims} \,
                      \mid \, \textit{var}
\end{align*}

@@

\section{Static analysis}

@@

>>>>

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{color}
\usepackage{graphics}
\usepackage{fancyvrb}
\usepackage{subfigure}

\CustomVerbatimEnvironment{SpecVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.65cm,
xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.97}
\CustomVerbatimEnvironment{ExmVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.65cm,
xrightmargin=0.2cm,baselinestretch=0.97,numbers=left}
\CustomVerbatimEnvironment{IVerbatim}{Verbatim}{fontsize=\relsize{-1},xleftmargin=0.65cm,
xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.97}

@@

\newcommand{\dimId}{\texttt{dim}}

@@

\title{Time and space specifications for stencils}
\authorinfo{}{}

@@

\emph{Stencils} are a ubiquitous programming pattern in many numerical
applications. Informally, a stencil computation computes
an array, where the value at each index $i$ of this array is
calculated from a set of neighbouring values to $i$ in some input
array(s), \eg{}, the Game of Life, convolutions in image processing, 
approximations to differential equations. For example, the following
psuedo-code iteratively applies the discrete Laplace
operator (an approximation to a derivative):
%
\begin{ExmVerbatim}
do iter = 0, itermax
   do i = 1, (n-1)
      b(i) = a(i-1) - 2*a(i) + a(i+1)
   a = b
\end{ExmVerbatim}
%
Line 3 is the core of the stencil computation, calculating
the value at \texttt{b[i]} from a neighbourhood of elements about
\texttt{a[i]}. Line 4 swaps 
\texttt{a} and \texttt{b} between iterations, where \texttt{b} becomes the
input for the next iteration.

%This stencil computation exhibits statically decideable
%spatial and temporal relationships between $\texttt{a}$ and
%$\texttt{b}$. 
In this simple example, the spatial and temporal relationship
between \texttt{a} and \texttt{b} is easy to understand. However,
more complex stencil computations can be much harder to understand
and subsequently more prone to error. For example, Figure~\ref{ref:navier-stokes-fragment}
shows a couple of lines from a Navier-Stokes fluid simulator in
which two arrays are read from with different data access patterns,
across two dimensions. The interaction is much harder to understand,
with the potential for the developer to accidentally introduce an
error, for example writing $\texttt{(i-1,j)}$ instead of
$\texttt{(i+1,j)}$.

We introduce a simple specification for stencil comptuations
that abstracts the fine grained detail of their access patterns.
In practise, most stencil computations have a regular shape that
can be described simply and abstractly. In the case of our
first Laplace example, our analysis tool infers the following specification:
%
\begin{SpecVerbatim}
a : centered depth=1 dims=1
b : backward depth=1\{a\} dims=t
\end{SpecVerbatim}
%
The first line explains that \texttt{a} is read from
with a symmetrical stencil pattern (``centered'') to a depth
of one in each direction in its first (and only) dimension.
The second line explains the temporal relationship between
\texttt{b} and \texttt{a}: that the previous time step for \texttt{b}
is actually provided by \texttt{a}. The dimension \texttt{t} is a
reserved dimension identifiers for temporal specifications. In
the case of the Navier-Stokes example, its inferred stencil is shown
in Figure~\ref{ref:navier-stokes-fragment}(b).

This paper makes the following contributions:
%
\begin{itemize}
\item introduces a specification language for
stencil computations that captures many common forms
of stencil (Section~\ref{sec:lang});
\item we detail specification inference and checking
algorithsm (Section~\ref{sec:analysis});
\item we evaluate our implementation of the approach
in the CamFort tool for Fortran verification, studying
a number of example programs to assess the usefulness
of this approach.
\dnote{insert results here}
\end{itemize}
%

\begin{figure}
\begin{ExmVerbatim}[firstnumber=20]
du2dx = ((u(i,j)+u(i+1,j))*(u(i,j)+u(i+1,j))+ 
    gamma*abs(u(i,j)+u(i+1,j))*(u(i,j)-u(i+1,j))- 
    (u(i-1,j)+u(i,j))*(u(i-1,j)+u(i,j))- 
    gamma*abs(u(i-1,j)+u(i,j))*(u(i-1,j)-u(i,j))) 
    /(4.0*delx)

duvdy = ((v(i,j)+v(i+1,j))*(u(i,j)+u(i,j+1))+ 
   gamma*abs(v(i,j)+v(i+1,j))*(u(i,j)-u(i,j+1))- 
   (v(i,j-1)+v(i+1,j-1))*(u(i,j-1)+u(i,j))-  
   gamma*abs(v(i,j-1)+v(i+1,j-1))*(u(i,j-1)-u(i,j))) 
   /(4.0*dely)

laplu = (u(i+1,j)-2.0*u(i,j)+u(i-1,j))/delx/delx+ 
          (u(i,j+1)-2.0*u(i,j)+u(i,j-1))/dely/dely
\end{ExmVerbatim} 
(a). Excerpt from the Fortran code, 
showing highly-detailed stencil computations. \\

\begin{SpecVerbatim}[xleftmargin=0.1cm]
20-24  u: centered depth=1 dim=1
26-30  u: centered depth=1 dim=2
       v: forward depth=1 dim=1, backward depth=1 dim=1
32-33  u: centered depth=1 dim=1,2
\end{SpecVerbatim}
(b). Inferred stencil specification from CamFort
\caption{Fragment of Navier-Stokes fluid simulator and its specification}
\label{ref:navier-stokes-fragment}
\end{figure}

\section{Stencil specification language}
\label{sec:lang}

\begin{figure}
\begin{align*}
spec ::= & \; S  \; \texttt{,} \; spec \, \mid \, S \\[0.5em]
%
S ::= &  \; \texttt{forward} \; [\texttt{depth=}\textit{depth}] \; \texttt{dims=}\textit{dims} \\
& \; \mid \texttt{backward} \; [\texttt{depth=}\textit{depth}] \; \texttt{dims=}\textit{dims} \\
& \; \mid \texttt{centered} \; [\texttt{depth=}\textit{depth}] \; \texttt{dims=}\textit{dims} \\
& \; \mid \texttt{unspecified} \; [\texttt{dims=}\textit{dims}] \\
& \; \mid \texttt{read-once} \; S \\[0.5em]
\textit{depth} ::= & \; \mathbb{N} \, \mid \, \textit{temp-spec} \\ 
\textit{temp-spec} ::= & \mathbb{N}\{var\} \\
\textit{dims}  ::= & \;\mathbb{N}_{>0} \; \texttt{,} \; \textit{dims} \,
                      \mid \, \mathbb{N}_{>0} \, \mid \, 
                     \texttt{t}
\end{align*}
\caption{Syntax of specifications}
\end{figure}

\begin{figure}
\begin{align*}
(\texttt{a} : \; \texttt{bwd} \; \texttt{depth=1\{b\}} \;
  \texttt{dim=t})
\, & \equiv \,
(\texttt{b} : \; \texttt{bwd} \; \texttt{depth=1\{a\}} \;
  \texttt{dim=t})
\\ 
(\texttt{a} : \; \texttt{fwd} \; \texttt{depth=1\{b\}} \;
  \texttt{dim=t})
\, & \equiv \,
(\texttt{b} : \; \texttt{fwd} \; \texttt{depth=1\{a\}} \;
  \texttt{dim=t})
\\
(a : \texttt{read-once} \; S) \; & \equiv \; (a : \texttt{read-once} \;\,
  \texttt{read-once} \; \, S)
\end{align*}
\caption{Equations on specifications}
\end{figure}

\subsection{Time specifications}

\begin{ExmVerbatim}
for t = 1, itermax:
   for i = 1, (n-1):
      a[t, i] = a[t-1, i-1] - 2*a[t-1, i] + a[t-1, i+1]
\end{ExmVerbatim}

\begin{SpecVerbatim}
a(t=0) : bwd, depth=1
a(1)   : centered, depth=1
\end{SpecVerbatim}

\begin{SpecVerbatim}
a : centered depth=1 dim=1
a : backward depth=1 dim=0
\end{SpecVerbatim}

\subsection{Safety and sub-specifications}

\begin{figure}[t]
{\framebox{
\begin{minipage}{0.88\linewidth}
\begin{align*}
\hspace{-0.7em}
\begin{array}{c}
\dfrac{}{S <: S}(\textsc{refl}) \qquad \dfrac{R <: S \quad S <: T}{R <:
  T}(\textsc{trans}) \\[1.5em]
\dfrac{}{S <: \; \texttt{unspecified dims=}ds}(\top) 
\\[1.5em]
\dfrac{}{\texttt{read-once} \, S <: S}(\textsc{rep})
\\[1.5em]
\setlength{\arraycolsep}{0.1em}
\dfrac{\hspace{3em} ds \subseteq es \; \wedge \; n \leq m \hspace{3em}}
{\begin{array}{rl}
\texttt{fwd} \; \texttt{depth=}n \; \texttt{dim=}ds & <: \texttt{fwd} \;
  \texttt{depth=}m \; \texttt{dim=}es 
%& \;\, ds \subseteq es \wedge n \leq m 
\\
\wedge \; \texttt{bwd} \; \texttt{depth=}n \; \texttt{dim=}ds & <: \texttt{bwd} \;
  \texttt{depth=}m \; \texttt{dim=}es 
%& \;\, ds \subseteq es \wedge n \leq m 
\\
\wedge \; \texttt{fwd} \; \texttt{depth=}n \; \texttt{dim=}ds & <: \texttt{sym} \;
  \texttt{depth=}m \; \texttt{dim=}es 
%& \;\, ds \subseteq es \wedge n \leq m 
\\
\wedge \; \texttt{bwd} \; \texttt{depth=}n \; \texttt{dim=}ds & <: \texttt{sym} \;
  \texttt{depth=}m \; \texttt{dim=}es 
\end{array}}{\footnotesize{(\textsc{cover})}}\hspace{-0.1em}
\end{array}
\end{align*}
\end{minipage}}}
\caption{Definition of sub-specification relation}
\end{figure}

@@

\section{Specification inference and checking}
\label{sec:analysis}

@@

\subsection{Infering temporal specifications}

The reserved dimension \dimId{t}
